Photo-D with LSST: stellar photometric distances out to the edge of the Galaxy

### 2024-05-28 paper/analysis.tex
comments from coauthors

Luminosity-color models introduced in \S\ref{sec:method} are expressed as lookup tables, rather than as analytic functions. They are initially defined
on a two-dimensional grid, and after accounting for dust extinction, on a three-dimensional grid. We use equidistant grids for all three
model parameters: 0.01 mag for $M_r$, 0.05 dex for $[Fe/H]$ and a range of steps from 0.005 mag to 0.02 mag for $A_r$, depending on its
prior range. These grids have sufficient resolution for numerical analysis of the posterior probability distributions and typically
result (depending on the maximum allowed $A_r$ value) in several million grid points. For a given star, exhaustive grid search method simply iterates
over all these points and for each computes likelihood and ultimately the posterior pdf.

Significant acceleration can be achieved by using a two-step iteration, with degraded grids in the first step, but still fine
enough to sufficiently resolve posterior pdf to approximately estimate its behavior around its maximum. We find that degrading $M_r$ step to 0.05 mag and
$[Fe/H]$ step to 0.1 dex works well in practice and results in about 10 times shorter run time. After the first iteration, the range $\pm 3
\sigma$ around the posterior  maximum, where $\sigma$ is standard deviation, is estimated using marginal distribution for each model
parameter. The calculation of posterior is then repeated
using the finest grid but only in the $\pm 3 \sigma$  neighborhood (cube) around posterior maximum, which includes 100 to 1000 times fewer grid locations.

@@

Markov Chain Monte Carlo methods have been successfully used in this context (e.g., \citealt{2014ApJ...783..114G}). 
Given that we aim to process LSST photometry for about 10$^{10}$ stars, it is worthwhile to compare MCMC runtime
to runtime for our accelerated exhaustive grid search method. We considered two implementations, based on {\it PyMC3} \citep{pymc2023} and
{\it emcee} \citep{2013PASP..125..306F}. The latter was about 10-20 times faster than the former; for example, with 4 ``walkers'' and 500
iterations, it took on average about 0.1 sec per star. Therefore, in this low-dimensional case (there are only three model parameters) the
grid search method described above is about an order of magnitude faster than {\it emcee} implementation.

@@

Further acceleration can be achieved using Neural Network method. In a companion paper by Mrakov\v{c}i\'{c} et al., we describe
how to use a small subsample of stars to train a neural network model and then estimate model parameters and their uncertainty for the remaining
stars using trained network. This approach yields runtimes of about 1 milisec per star, that is, shorter by an order of magnitude than when
the grid search method is used for the full sample. With this additional speed up, a sample of 10 billion stars can be processed in
about 10 hours using a 300-core machine and pipeline described in the next Section.

@@

Our simulated catalogs are based on TRILEGAL simulations of LSST stellar content by \cite{2022ApJS..262...22D}, already
mentioned in the context of priors in \S\ref{sec:priors}. We use these simulations to generate distributions in the
$M_r$--$[Fe/H]$--$A_r$--magnitude--sky position space. Given these quantities for a sample of simulated TRILEGAL stars,
we generate simulated LSST photometry for each star as follows: 
\begin{enumerate}
\item
Given relevant model parameters (e.g., $M_r$ and $[Fe/H]$ for main sequence and red giant stars, $M_r$ and log(g) for
white dwarfs, the total system luminosity and the component luminosity ratio in the $r$ band for unresolved binaries),
we use luminosity-color tracks (see Figure~\ref{fig:augmLocus}) to
assign photometric noise-free colors. We do not use original TRILEGAL
colors because for proper statistical tests colors must be consistent
with luminosity-color tracks used for likelihood computations. 
\item
Using apparent $r$ band magnitude from TRILEGAL simulation, corresponding to simulated distance and $M_r$, and 
colors from the previous step, we generate magnitudes in all the remaining bands.
\item
We generate per-band photometric uncertainties, $\sigma_b$, using eqs. 4 and 5 from \cite{2019ApJ...873..111I} and
per-band 5-$\sigma$ depths (the so-called $m_5$) expected for co-added LSST data from \cite{2022ApJS..258....1B}.
\item
We draw Gaussian noise from N(0, $\sigma_b$) and add it to each magnitude to obtain ``observed'' magnitudes.   
\item
Finally, we recompute photometric uncertainties to be treated as ``observational'' uncertainties using ``observed'' magnitudes
(because the first set of photometric uncertainties are derived using true noise-free magnitudes). 
\end{enumerate}

For numerical tests, we select a region from the so-called SDSS Stripe 82, defined by $340^{\deg} < {\rm R.A.} < 350^{\deg}$ and
$-1.3^{\deg} < \delta < 1.3^{\deg}$, which contains 280,000 simulated main sequence and red giants stars that satisfy
$r<26$ and $u<27$. Their distributions of model parameters $M_r$, $[Fe/H]$, and interstellar dust extinction along the line of sight in
the $r$ band, $A_r$,  in the $u$ vs. $g-i$ color-magnitude diagram are shown in Figure~\ref{fig:qpBmeans}.

\begin{figure*}[ht!]
\hskip -0.3in
\includegraphics[width=1.08\textwidth,angle=0]{figures/qpBmeans_SDSSpatchRA340-350-simLSST_FeH.png}
\vskip -0.3in
\caption{The mean per-pixel values of absolute magnitude $M_r$, metallicity $[Fe/H]$, and interstellar dust extinction along the
  line of sight in the $r$ band, $A_r$, for a TRILEGAL-simulated sample of 280,000 main
  sequence and red giants stars with $r<26$, $u<27$, $340^{\degree} < {\rm R.A.} < 350^{\degree}$
  and $-1.3^{\degree} < \delta < 1.3^{\degree}$ (a small patch from the SDSS Stripe 82 region) in the $u$ vs. $g-i$
  color-magnitude diagram. The mean values are color-coded according to legend below each panel. Contours
  visualize the sample distribution in each diagram. The strong variation of $M_r$ with the $g-i$ color is seen in the left panel,
  except for the small patch with $u<16.5$ and $1.0 < g-i < 1.5$ where red giant stars dominate. The
  diagonal iso-metallicity boundaries in the middle panel closely correspond to distance
  (in the range from about 1 kpc to about 100 kpc in the lower left corner). The values of
  dust extinction, shown in the right panel, are not large ($<0.2$) because the
  selected field is at high Galactic latitudes (centered on $b = -52^{\degree}$).}
\label{fig:qpBmeans}
\end{figure*}

Figures~\ref{fig:perfVSestParamsMr} and \ref{fig:perfVSestParamsFeH} show summarized statistical performance of the Bayesian
method (bias, scatter, $\chi^2$) in the $u$ vs. $g-i$ color-magnitude diagram. In this test based on a high-galactic
latitude field, we assumed that all stars are beyond the dust layer and fixed $A_r$ to its known true value (in practice provided by dust
maps such as those from \citealt{schlegel_maps_1998}). When $A_r$ is considered as a free parameter in sky regions with small $A_r$,
the best-fit $A_r$ values are often underestimated for stars in the blue part of the stellar locus (because the reddening vector and locus
are nearly parallel; for more details, see \S2.7.1 in \citealt{2012ApJ...757..166B}). We note that at faint magnitudes probed by LSST, this
assumption will only break down very close to the Galactic plane.

@@

Analysis of the method's performance in the preceding subsection was based on a simulated sample with ``observed'' colors generated using
exactly the same luminosity-color model tracks as those used in fitting. Therefore, that analysis cannot test validity of those tracks in an absolute sense.
While these empirical luminosity-color model tracks were derived using SDSS observations of
globular clusters with known distances and metallicities, here we validate them further using trigonometric distances recently obtained by Gaia.

For validation, we used SDSS Stripe 82 Standard Star Catalog, recalibrated by \cite{2021MNRAS.505.5941T},
which provides the most precise available photometry in SDSS bands (due to averaging of many repeated SDSS observations). 
For stars listed in the catalog, we extracted Gaia measurements and ``photogeometric'' distances from 
\cite{bailer-jones_estimating_2021}. Out of 841,000 stars listed in the Standard Star Catalog, there are 415,000 stars with $r < 22$, $u<22$ and a Gaia
match within 0.15 arcsec (after correcting for proper motion using Gaia measurements). Their distribution in color-color and
color-magnitude diagrams is shown in Figures~\ref{fig:3dataDiags} and \ref{fig:3HRdiags}. 
We estimated $M_r$ and $[Fe/H]$ for these stars used the same procedure and assumptions as for the simulated sample from the
preceding subsection. In particular, we assumed that all stars are beyond the dust layer and fixed $A_r$ to its value provided by the dust
maps from \cite{schlegel_maps_1998}.

@@

The median contribution of Gaia's parallax measurement uncertainty to Gaia's absolute
magnitude uncertainty for this subsample is about 0.12 mag. This estimate implies that the median uncertainty of SDSS-based $M_r$
estimates is 0.27 mag (corresponding to about 14\% distance uncertainty). On the other hand, when only stars with signal-to-noise
ratio for Gaia's trigonometric parallax measurement above 50 are considered, the implied uncertainty of SDSS-based $M_r$
estimates is 0.21 mag (corresponding to about 10\% distance uncertainty).  Recalling that the statistical uncertainties for $M_r$
estimates using simulated sample in the preceding section are 0.10 mag at the bright end, these somewhat larger uncertainties may
imply additional statistical effects not accounted for in our analysis, and/or imperfect luminosity color-tracks.
Figure~\ref{fig:gaiasdss1} demonstrates that there are no concerningly large systematic errors in photometric $M_r$ estimates with
respect to photometric noise ($u$ band magnitude), $g-r$ color and photometric metallicity estimate (though note that the metallicity
range corresponds only to disk stars).

We extended our analysis to fainter magnitude limit by replacing Gaia's magnitudes based on trigonometric distance with those based on
the so-called ``photogeometric'' distances from \cite{bailer-jones_estimating_2021}. This comparison goes about 2
magnitudes deeper and also extends to low halo-like metallicity range. The scatter between photometric $M_r$ based on SDSS data and
$M_r$ based on ``photogeometric'' distances is 0.38 mag. As shown in Figure~\ref{fig:gaiasdss2}, the systematic error
increases to about 0.2-0.3 at the blue and low-metallicity edge of the stellar locus. When stars with $0.2 < g-r < 0.3$ are excluded,
systematic errors at the low-metallicity ($[Fe/H]<-1$) end disappear, indicating that it is the $M_r$ as a function of color relation that fails at the blue end,
rather than the shift of $M_r$ as a function of $[Fe/H]$. Such edge effects will have to be recalibrated when LSST photometry becomes available
(both using Gaia data and globular clusters).

>>>>

Luminosity-color models introduced in \S\ref{sec:method} are expressed as lookup tables, rather than as analytic functions. They are initially defined
on a two-dimensional grid, and after accounting for dust extinction, on a three-dimensional grid. We use equidistant grids for all three
model parameters: 0.01 mag for $M_r$, 0.05 dex for $[Fe/H]$ and a range of steps from 0.005 mag to 0.02 mag for $A_r$, depending on its
prior range. These grids have sufficient resolution for numerical analysis of the posterior probability distributions and typically
result (depending on the maximum allowed $A_r$ value) in several
million grid points. For a given star, the exhaustive grid search method simply iterates
over all these points and for each computes likelihood and ultimately the posterior pdf.

Significant acceleration can be achieved by using a two-step iteration, with degraded grids in the first step, but still fine
enough to sufficiently resolve the posterior pdf to approximately estimate its behavior around its maximum. We find that degrading the
$M_r$ step to 0.05 mag and the $[Fe/H]$ step to 0.1 dex works well in practice and results in a about ten times shorter run time. After the first iteration, the range $\pm 3
\sigma$ around the posterior  maximum, where $\sigma$ is standard deviation, is estimated using the marginal distribution for each model
parameter. The calculation of the posterior is then repeated using the finest grid but only in the $\pm 3 \sigma$  neighborhood
(cube) around the posterior maximum, which includes 100 to 1000 times fewer grid locations.

@@

Markov Chain Monte Carlo methods have been successfully used in this context (e.g., \citealt{2014ApJ...783..114G}). 
Given that we aim to process LSST photometry for about 10$^{10}$ stars, it is worthwhile to compare the MCMC runtime
to the runtime for our accelerated exhaustive grid search method. We considered two implementations, based on {\it PyMC3} \citep{pymc2023} and
{\it emcee} \citep{2013PASP..125..306F}. The latter was about 10-20 times faster than the former; for example, with 4 ``walkers'' and 500
iterations, it took on average about 0.1 sec per star. Therefore, in this low-dimensional case (there are only three model parameters) the
grid search method described above is about an order of magnitude faster than the {\it emcee} implementation.

@@

Further acceleration can be achieved using the Neural Network method. In a companion paper by Mrakov\v{c}i\'{c} et al., we describe
how to use a small subsample of stars to train a neural network model and then estimate model parameters and their uncertainty for the remaining
stars using the trained network. This approach yields runtimes of about 1 milisec per star, that is, shorter by an order of magnitude than when
the grid search method is used for the full sample. With this additional speed up, a sample of 10 billion stars can be processed in
about 10 hours using a 300-core machine and the pipeline described in the next Section.

@@

Our simulated catalogs are based on TRILEGAL simulations of LSST stellar content by \cite{2022ApJS..262...22D}, already
mentioned in the context of priors in \S\ref{sec:priors}. We use these simulations to generate distributions in the
$M_r$--$[Fe/H]$--$A_r$--magnitude--sky position space. Given these quantities for a sample of simulated TRILEGAL stars,
we generate simulated LSST photometry for each star as follows: 
\begin{enumerate}
\item
Given relevant model parameters (e.g., $M_r$ and $[Fe/H]$ for main sequence and red giant stars, $M_r$ and log(g) for
white dwarfs, the total system luminosity and the component luminosity ratio in the $r$ band for unresolved binaries),
we use luminosity-color tracks (see Figure~\ref{fig:augmLocus}) to
assign photometric noise-free colors. We do not use the original TRILEGAL
colors because for proper statistical tests colors must be consistent
with luminosity-color tracks used for likelihood computations. 
\item
Using the apparent $r$ band magnitude from the TRILEGAL simulation, corresponding to simulated distance and $M_r$, and 
colors from the previous step, we generate magnitudes in all the remaining bands.
\item
We generate per-band photometric uncertainties, $\sigma_b$, using eqs. 4 and 5 from \cite{2019ApJ...873..111I} and
per-band 5-$\sigma$ depths (the so-called $m_5$) expected for co-added LSST data from \cite{2022ApJS..258....1B}.
\item
We draw Gaussian noise from N(0, $\sigma_b$) and add it to each magnitude to obtain ``observed'' magnitudes.   
\item
Finally, we recompute photometric uncertainties to be treated as ``observational'' uncertainties using ``observed'' magnitudes
(because the first set of photometric uncertainties are derived using true noise-free magnitudes). 
\end{enumerate}

For numerical tests, we select a region from the so-called SDSS Stripe 82, defined by $340^{\degree} < {\rm R.A.} < 350^{\degree}$ and
$-1.3^{\degree} < \delta < 1.3^{\degree}$, which contains 280,000 simulated main sequence and red giants stars that satisfy
$r<26$ and $u<27$. Their distributions of model parameters $M_r$, $[Fe/H]$, and interstellar dust extinction along the line of sight in
the $r$ band, $A_r$,  in the $u$ vs. $g-i$ color-magnitude diagram are shown in Figure~\ref{fig:qpBmeans}.

\begin{figure*}[ht!]
\hskip -0.3in
\includegraphics[width=1.08\textwidth,angle=0]{figures/qpBmeans_SDSSpatchRA340-350-simLSST_FeH.png}
\vskip -0.3in
\caption{The mean per-pixel values of absolute magnitude $M_r$, metallicity $[Fe/H]$, and interstellar dust extinction along the
  line of sight in the $r$ band, $A_r$, for a TRILEGAL-simulated sample of 280,000 main
  sequence and red giants stars with $r<26$, $u<27$, $340^{\degree} < {\rm R.A.} < 350^{\degree}$
  and $-1.3^{\degree} < \delta < 1.3^{\degree}$ (a small patch from the SDSS Stripe 82 region) in the $u$ vs. $g-i$
  color-magnitude diagram. The mean values are color-coded according to the legend below each panel. Contours
  visualize the sample distribution in each diagram. The strong variation of $M_r$ with the $g-i$ color is seen in the left panel,
  except for the small patch with $u<16.5$ and $1.0 < g-i < 1.5$ where red giant stars dominate. The
  diagonal iso-metallicity boundaries in the middle panel closely correspond to distance
  (in the range from about 1 kpc to about 100 kpc in the lower left corner). The values of
  dust extinction, shown in the right panel, are not large ($<0.2$) because the
  selected field is at high Galactic latitudes (centered on $b = -52^{\degree}$).}
\label{fig:qpBmeans}
\end{figure*}

Figures~\ref{fig:perfVSestParamsMr} and \ref{fig:perfVSestParamsFeH} show the summarized statistical performance of the Bayesian
method (bias, scatter, $\chi^2$) in the $u$ vs. $g-i$ color-magnitude diagram. In this test based on a high-galactic
latitude field, we assumed that all stars are beyond the dust layer and fixed $A_r$ to its known true value (in practice provided by dust
maps such as those from \citealt{schlegel_maps_1998}). When $A_r$ is considered as a free parameter in sky regions with small $A_r$,
the best-fit $A_r$ values are often underestimated for stars in the blue part of the stellar locus (because the reddening vector and locus
are nearly parallel; for more details, see \S2.7.1 in \citealt{2012ApJ...757..166B}). We note that at faint magnitudes probed by LSST, this
assumption will only break down very close to the Galactic plane.

@@

Analysis of the method's performance in the preceding subsection was based on a simulated sample with ``observed'' colors generated using
exactly the same luminosity-color model tracks as those used in fitting. Therefore, that analysis cannot test the validity of those tracks in an absolute sense.
While these empirical luminosity-color model tracks were derived using SDSS observations of
globular clusters with known distances and metallicities, here we validate them further using trigonometric distances recently obtained by Gaia.

For validation, we used the SDSS Stripe 82 Standard Star Catalog, recalibrated by \cite{2021MNRAS.505.5941T},
which provides the most precise available photometry in SDSS bands (due to averaging of many repeated SDSS observations). 
For stars listed in the catalog, we extracted Gaia measurements and ``photogeometric'' distances from 
\cite{bailer-jones_estimating_2021}. Out of 841,000 stars listed in the Standard Star Catalog, there are 415,000 stars with $r < 22$, $u<22$ and a Gaia
match within 0.15 arcsec (after correcting for proper motion using Gaia measurements). Their distribution in color-color and
color-magnitude diagrams is shown in Figures~\ref{fig:3dataDiags} and \ref{fig:3HRdiags}. 
We estimated $M_r$ and $[Fe/H]$ for these stars using the same procedure and assumptions as for the simulated sample from the
preceding subsection. In particular, we assumed that all stars are beyond the dust layer and fixed $A_r$ to its value provided by the dust
maps from \cite{schlegel_maps_1998}.

@@

The median contribution of Gaia's parallax measurement uncertainty to Gaia's absolute
magnitude uncertainty for this subsample is about 0.12 mag. This estimate implies that the median uncertainty of SDSS-based $M_r$
estimates is 0.27 mag (corresponding to about 14\% distance uncertainty). On the other hand, when only stars with signal-to-noise
ratio for Gaia's trigonometric parallax measurement above 50 are considered, the implied uncertainty of SDSS-based $M_r$
estimates is 0.21 mag (corresponding to about 10\% distance uncertainty).  Recalling that the statistical uncertainties for $M_r$
estimates using simulated sample in the preceding section are 0.10 mag at the bright end, these somewhat larger uncertainties may
imply additional statistical effects not accounted for in our analysis, and/or imperfect luminosity color-tracks.
Figure~\ref{fig:gaiasdss1} demonstrates that there are no concerningly large systematic errors in photometric $M_r$ estimates with
respect to photometric noise ($u$ band magnitude), $g-r$ color and photometric metallicity estimates (though note that the metallicity
range corresponds only to disk stars).

We extended our analysis to fainter magnitude limits by replacing Gaia's magnitudes based on trigonometric distance with those based on
the so-called ``photogeometric'' distances from \cite{bailer-jones_estimating_2021}. This comparison goes about 2
magnitudes deeper and also extends to low halo-like metallicity range. The scatter between photometric $M_r$ based on SDSS data and
$M_r$ based on ``photogeometric'' distances is 0.38 mag. As shown in Figure~\ref{fig:gaiasdss2}, the systematic error
increases to about 0.2-0.3 at the blue and low-metallicity edge of the stellar locus. When stars with $0.2 < g-r < 0.3$ are excluded,
systematic errors at the low-metallicity ($[Fe/H]<-1$) end disappear, indicating that it is the $M_r$ as a function of color relation that fails at the blue end,
rather than the shift of $M_r$ as a function of $[Fe/H]$. Such edge effects will have to be recalibrated when LSST photometry becomes available
(both using Gaia data and globular clusters).
